---
title: "All Ipo"
author: "Kelly Endres and Connor Morozumi"
date: "April 17, 2020"
output: html_document
---
This R Markdown file contains all code for Endres et al. 2021 manuscript investigating the effects of drought on Ipomopsis aggregata niche breadth. We utilized Ipomopsis visitation data from several sites over five years, 1997, 1998, 2001, 2012, and 2018, with 2012 and 2018 determined to be drought years. Between years we compared visitor abundance, community change, and richness. We also analyzed weather data to confirm drought year determination. 

###Code Overview

1. Drought Analysis
+ Adonis
+ NMDS
2. Visitation Summary Statistics
3. Visitor Abundance
+ glmm abundance analysis
4. Richness
+ glmm analysis - Visit richness
+ iNEXT analysis - Visit richness
5. Community Change
+ Adonis analysis
+ NMDS



####Package Load and Data Import

```{r packages, include=FALSE}

#load packages 
library(glmmTMB)
library(fitdistrplus)
library(DHARMa)
library(ggplot2)
library(tidyverse)
library(vegan)
library(dplyr)
library(GGally)
library(reshape2)
library(iNEXT)
library(gridExtra)
library(viridis)
library(cowplot)
```

These analyses require the associated datasets:
gothicweather.csv
total visits per site per year by each visitor catagory.csv
visit.abundance.by.observation.day.csv

```{r load data}
weatherdf = read.csv("gothicweather.csv")

all= read.csv("total visits per site per year by each visitor catagory.csv")

all.years.ab= read.csv("visit.abundance.by.observation.day.csv")

```


### 1. Drought Analysis

Here we confirm that our catagorization of 2012 and 2018 as drought years is accurate 

```{r data import}

#import data
weatherdf<- read.csv("gothicweather.csv")

# water is in inches so convert to cm
weatherdf$Watercm= weatherdf$Water * 2.54

# remove winters before 1994/95
weatherdf = weatherdf[c(17:32,37:43),]

# if only 2012 and 2018 are classified as drought
weatherdf= weatherdf %>%
  mutate(Drought_New = as.factor(ifelse(Winter == "2011-12" | Winter== "2017-18", "Drought", "Non Drought")))


```

With the data loaded, it needs to be reformatted as a matrix. The non-numeric columns need to be removed, as they can't be used in the nmds. 

```{r matrix formatting}

#create matrix
mat <- as.matrix(weatherdf)

#rename rows
row.names(mat) <- weatherdf$Winter

#delete non-numeric columns
colnames(mat)[1] <- "1"
colnames(mat)[5] <- "5"
colnames(mat)[6] <- "6"
mat <- mat[, - c(1, 5, 6)]
```

The matrix needs to be formatted further to fit the form for the NMDS analysis. The drought column needs to be removed from the dataframe containing the rest of the data and put into its own value. 

```{r create df}

# Make dataframe
dmat <- as.data.frame(mat) 

#change columns to numeric
dmat$Total <- as.numeric(as.character(dmat$Total))
dmat$Apr.1 <- as.numeric(as.character(dmat$Apr.1))
dmat$May.1 <- as.numeric(as.character(dmat$May.1))
dmat$Water <- as.numeric(as.character(dmat$Water))
dmat$Above.freezing <- as.numeric(as.character(dmat$Above.freezing))
dmat$Days.covered <- as.numeric(as.character(dmat$Days.covered))
dmat$Drought <- as.numeric(as.character(dmat$Drought))

#create value with drought information
gr <- c(dmat$Drought)

#remove drought column from other data
colnames(dmat)[7] <- "7" #rename
dmat2 <- dmat[, -c(7:9)] #remove
```

Now that the data is formatted properly analysis can begin. We used analysis of variance with ADONIS and NMDS to visualize differences between drought and non-drought years. We expect that 2012 and 2018 will be drought years. 

# Adonis

Create a separate dataframe is of just Drought information. 


```{r}

des= weatherdf[,c(1, 12)] #remove everything but drought status

rownames(des) <- des[,1]
des[,1] <- NULL

```

Compute distances with Morisita-Horn

```{r}
dist.mat <- vegdist(dmat2, method = "horn")
```

Run analysis of variance with adonis2
```{r}

model= adonis2(dist.mat ~ Drought_New, des, perm = 999)

```

Drought is highly significant! Our drought categorization explains multivariate composites of many weather variables.

# NMDS
Visualize with the function metaMDS. 

```{r NMDS}

# run a NMDS ordination
ord <- metaMDS(dmat2)
```

Because we have a large dataset, the function automatically performs a square root transformation using Wisconsin standardization. 

Fit an envfit of apriori drought status
```{r}
data.envfit <- envfit(ord, gr, lwd = 2)
```


Plot the NMDS 

```{r plotting}

plot(ord, disp="sites", type="n")
  ordihull(ord, gr, col=c("mediumturquoise", "indianred1"), lwd=3)
  ordiellipse(ord, gr, draw="polygon", col=c("mediumturquoise", "indianred1"), alpha = c(50, 50))
  orditorp(ord,display="species",col="black",air=0.01, cex = .8)
  # orditorp(ord,display="sites", select = sel, col= "black", air=0.01,cex=1)
  points(ord, display = "sites", cex = 0.5, pch = 3, col = "black")
plot(data.envfit, col = "black", cex = .5)
```

This visualization gives several important pieces of information. Green/blue specifies Non-drought and red is Drought. We see that the two areas show no overlap, indicating that the years are different in terms of weather. 

### 2. Visitation Summary Statistics


```{r reshape visitation data for Table 1}
all.long= all %>% 
  pivot_longer(cols = Butterfly:Wasp ,names_to = "Species" , values_to = "total.visits")

Table1= merge(all.long %>% 
  group_by(Year, Species) %>% 
  summarise_at(vars(total.visits), funs(sum)) %>% 
  pivot_wider(names_from = Species, values_from = total.visits ),

  all.long %>% 
  group_by(Year) %>% 
  summarise_at(vars(total.visits), funs(sum)), by = "Year")

Table1 %>%
    mutate_at(vars(c("Bumblebee", "Butterfly","Hoverfly","Hummingbird","Other.fly","Solitary.bee","Wasp")), 
               function (x) (x/Table1$total.visits)*100)

```

## 3. Visitor Abundance
+ glmm abundance analysis

With the following code we compare visitor abundance between years. This is important as we are not able to look at straight abundances between years due to sampling effort differences. We want to compare these as it is an important aspect of how drought may impact plant-pollinator communities. 

To examine abundance differences and account for sampling effort differences, we will be using a linear mixed effect model, glmmTMB adding several different fixed and random effects. 


###Abundance Analysis

The first step to examine visitor abundance between drought and non-drought years is to determine the correct distribution for our data. It is most likely negative binomial, so we start by checking for that distribution.

```{r abundance distribution fit}

#fit data distribution
distcheck <- fitdist(all.years.ab$Visitor.Abundance, "nbinom")

#summarize nbinom fit
summary(distcheck)

#plot nbinom fit
plot(distcheck)
```

According the the plot of the distribution the data fits a negative binomial distribution. 

Now we are able to run our linear mixed effects model (glmmTMB) using a negative binomial distribution (nbinom2). Our response variable is given as total visitor abundance for each individual site visit. Drought, observer number, and amount of time observing are fixed effects, while site nested within year is a random effect. 

```{r abundance model}

#create model
abundancemod <- glmmTMB(Visitor.Abundance ~ Drought + Observers + Time + (1|Site/Year), all.years.ab, family = "nbinom2")

#summary
summary(abundancemod)
```

This indicates that the presence of drought impacts observed visitor abundance. The amount of time spent observing also impacts visitor abundance. 

Checking assumptions with the package `DHARMa`
```{r check model assumptions}
gg <- residuals(abundancemod)

plot(gg)

all.years.ab$residuals <- gg

all.years.ab$fitted <- abundancemod$fitted

simoutput <- simulateResiduals(fittedModel = abundancemod, n = 250)

plot(simoutput)
```

Plot 1a
```{r plot visit abundance}
# change factor levels for color plotting
all.years.ab$Drought <- factor(all.years.ab$Drought, levels = c("Non-drought", "Drought"))

visit=ggplot(all.years.ab, aes(x=Drought, y=Visitor.Abundance)) +
  geom_boxplot(outlier.alpha  =  0, size = 0.5, width = 0.5) +
  geom_point(aes(colour = Drought), alpha = 0.5, size = 3, position = position_jitter(w=0.2, h=0)) +
  theme_classic() + 
  labs(y="Visitor Abundance") +
  theme(legend.position = "none", axis.title.x = element_blank(), axis.title = element_text(size=16), axis.text=element_text(size=14), plot.title  = element_text(hjust = 0.02, face = "bold", size= 18)) +
  scale_color_viridis(discrete=TRUE)

visit
```
** note plot may look slightly different each time the code is run because there is a `geom_jitter` function that operates randomly. This makes only superficial difference and does not change the interpretation of the graph.


## 4. Richness

# Summary stats richness

```{r}
# Just averages across year-site 
all.years.ab %>% 
  group_by(Year, Site) %>%
  tally(Richness)
  
all.years.ab %>% 
  group_by(Drought) %>%
  summarize(mean_rich= mean(Richness))

# Mean richness was1.750 in drought years and 0.697 in non-drought years

# but this doesn't account for effort so mean of a mean of a mean might yield different results

# first average within site-year
mean= all.years.ab %>%
  group_by(Year, Site, Drought) %>%
  summarize(site_rich= mean(Richness))


year_mean= mean %>%
  group_by(Year, Drought) %>%
  summarize(year_rich= mean(site_rich))

year_mean %>%
  group_by(Drought) %>%
  summarize(grand_mean= mean(year_rich))

# changes values slightly
# Non-drought	0.6931687			
# Drought	1.4957265	

# Abundance the same way
# Just averages across year-site 
all.years %>% #CM CHANGE this to fit the name of the df in this markdown
  group_by(Year, Site) %>%
  tally(Visitor.Abundance)
  
all.years %>% 
  group_by(Drought) %>%
  summarize(mean_abund= mean(Visitor.Abundance))

# Mean abundance was 3.875 in drought years and 1.725 in non-drought years

# but this doesn't account for effort so mean of a mean of a mean might yield different results

# first average within site-year
mean= all.years.ab %>%
  group_by(Year, Site, Drought) %>%
  summarize(site_abud= mean(Visitor.Abundance))


year_mean= mean %>%
  group_by(Year, Drought) %>%
  summarize(year_abund= mean(site_abud))

year_mean %>%
  group_by(Drought) %>%
  summarize(grand_mean= mean(year_abund))

# changes values slightly
# Non-drought	1.670445			
# Drought	3.183226	
```

##Richness analysis with glmm

```{r}
c <- fitdist(all.years.ab$Richness, "nbinom")

plot(c)

model3 <- glmmTMB(Richness ~ Drought + Observers + Time + (1|Site/Year), all.years.ab, family = "nbinom2")

summary(model3)

# check residuals
gg2 <- residuals(model3)
plot(gg2)


all.years.ab$residuals2 <- gg2

all.years.ab$fitted2 <- model3$fitted

simoutput2 <- simulateResiduals(fittedModel = model3, n = 250)
plot(simoutput2)

richness= ggplot(all.years.ab, aes(x=Drought, y=Richness)) +
  geom_boxplot(outlier.alpha  =  0, size = 0.5, width = 0.5) +
  geom_point(aes(color = Drought), alpha = 0.3, size = 3, position = position_jitter(w=0.2, h=0)) +
  theme_classic() + 
  labs(x="Drought", y="Richness") +
  theme(legend.position = "none", axis.title.x = element_blank(), axis.title = element_text(size=16), axis.text=element_text(size=14), plot.title  = element_text(hjust = 0.02, face = "bold", size= 18)) +
  scale_color_viridis(discrete=TRUE)


```

Plot 1a and 1b in panel
```{r}
# use cowplot
panel=plot_grid(visit, richness, labels = c('A', 'B'))

ggsave("Figure 1 Abundance and Richness_revision", panel, device="pdf", width= 16.6, height = 14.0, units= "cm")
```

# iNEXT analysis - Visit richness

```{r}
totals= as.data.frame(t(Table1))
names(totals) <- totals[1,]
totals <- totals[-1,]

#remove totals row
totals=totals[-8,]

# change to a matrix
tv = as.matrix(totals)
```

Next the drought and non-drought data is combined into a different dataframe. Sum the drought data (2012 and 2018) and then the non-drought data (1997, 1998, 2001). 

```{r}
grouped=totals %>% 
  transmute(Drought= `2012`+ `2018`, Nondrought= `1997`+ `1998` +`2001`)
```


Next is the actual iNEXT analysis. We use only hill numbers 0 (species richness) 

#### by year

separate the years by diversity order (here just species richness). 

```{r inext by year}
#species richness
d1 <- iNEXT(tv, q=c(0), datatype="abundance", endpoint=NULL)

```


```{r plot 2b}

# plot richness

g3 <- ggiNEXT(d1, type=1, grey=T) + 
  theme_classic() + 
  xlim(0,350) +
  ylim(0, 7.5) +
  scale_colour_manual(values=c("#440154FF", "#440154FF", "#440154FF","#FDE725FF","#FDE725FF")) +
  scale_fill_manual(values=c("#440154FF", "#440154FF", "#440154FF","#FDE725FF","#FDE725FF")) +
  scale_shape_manual(values=c(15, 16, 17, 19, 18)) +
  labs(x = "Number of individual visits", y = "Visitor category diversity") +
  theme(axis.title = element_text(size=16), axis.text=element_text(size=14), plot.title = element_text(hjust = 0.02, size= 18, face="bold"), legend.title=element_blank(),legend.text = element_text(size=8), legend.position = c(0.89,0.3)) + 
 guides(linetype = FALSE)

# change size of text
gb_3 <- ggplot_build(g3)
gb_3$data[[1]]$size <- 4
gt3 <- ggplot_gtable(gb_3)

```

# By Drought vs non-drought

```{r grouped inext}
i1 <- iNEXT(grouped, q=c(0), datatype="abundance", endpoint=NULL)
```

```{r plot fig 2}
# richness
gt4 <- ggiNEXT(i1, type=1, grey=T) + 
  xlim(0,300) +
  ylim(0, 7.5) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.02, face = "bold", size= 18), axis.title = element_text(size=16), axis.text=element_text(size=14), legend.position = c(0.85,0.2), legend.title=element_blank()) + 
 guides(linetype = FALSE) +
  scale_colour_manual(values=c("#FDE725FF","#440154FF")) +
  scale_fill_manual(values=c( "#FDE725FF","#440154FF")) +
  labs(x = "Number of individual visits", y = "Visitor category diversity") 
   # labs(title = "(a)")

gb_4 <- ggplot_build(gt4)
gb_4$data[[1]]$size <- 4
gt4 <- ggplot_gtable(gb_4)



#graph together

panel= grid.arrange(gt4, gt3, ncol=2)
ggsave("Figure 2 Rarefied Richness.pdf", panel, device="pdf", width= 16.6, height = 12, units= "cm")
```

### 5. Community Change
+ Adonis analysis
+ NMDS

This set of analyses looks at differences in community composition between drought and non-drought years via dissimilarity using permutational multivariate analyses. This requires construction  of community similarity/dissimilarity matrices and conducting Adonis analyses, with drought vs. nondrought as the fixed effects. Adonis is useful for this analysis as it uses strata which takes into account random effects. We use the Morisita-Horn index with Adonis as it is insensitive to differences in sampling effort among categories. Therefore we can use raw data without correcting for sampling effort. 

####Data formatting

Load dataframe containing total visits per site by each visitor category. 

Make df into a matrix
``` {r}
all= column_to_rownames(all, "Site")
#remove descriptive columns
all.years <- all[,-c(8, 9)]
#make all column numeric
all.years[] <- lapply(all.years, as.numeric)
#transform into matrix
all.years <- as.matrix(all.years)
```

To go with this dataframe we are creating a second dataframe that contains site, year, and drought information. 

```{r community year other variables dataframe}
#create descriptive dataframe
all.des <- as.data.frame(all[,-c(1:7)])

#create value with drought information
Drought <- c(all$Drought)
```

####Adonis analyses

Now that the data formatting is complete community composition analyses can be completed. Here we are using the Morisita-Horn method (horn) to examine visitation between years. 

```{r adonis analyses}

# adonis(beta ~ Drought, all.des, perm = 999)
adonis(all.years ~ Drought, all.des, perm = 100000, method= "horn")
```

Looking at the output of the permutational multivariate analysis, we do not find statistically significant differences in community composition (p > 0.05). 


####NMDS analysis
```{r NMDS}
# run NMDS ordination
nmds <- metaMDS(all.years, distance = "horn")
```


Analyzing code outputs:

```{r analyze output} 
#call function output
nmds
```

This gives us lots of information. We can see the transformations used on the data (wisconsin(sqrt)), the distance measure used (horn), number of dimensions (2), as well as other outputs. The output is automatically rotated so that axis 1 has larger explanatory power. Goodness of fit is given with stress. Here, stress = 0.18 indicating that the output is a fair match for this dataset. 

Plot it
```{r plot fig3}

#plot NMDS
  pdf('fig3.pdf')
  pl <- ordiplot(nmds, type = "none")
points(pl, "sites", pch=c(rep(17, 9), rep(19, 33)), col=c(rep("#440154FF", 9), rep("#FDE725FF", 33)), bg=c(rep("#440154FF", 9), rep("#FDE725FF", 33)), cex=1.2)
ordiellipse(nmds,groups=Drought,draw="polygon",col=c("#FDE725FF", "#440154FF"),label=F)
text(pl, "species", col="black", cex=1.2, font = 1)
dev.off()

```


 We see that the two areas show some overlap indicating partial similarities in composition confirmed by adonis. 




